EEG characterization and classification based on histograms of visual gradients

### 2017-09-17 articlebackup.tex
Submitted version

%=================================================================
% Full title of the paper (Capitalized)
\Title{EEG characterization and classification based on Histogram of Gradients of Signal Plots}

% Authors, for the paper (add full first names)
\Author{Rodrigo Ramele $^{1,\dagger}$*, Ana Julia Villar $^{1}$ and Juan Miguel Santos  $^{1}$}
% Authors, for metadata in PDF
\AuthorNames{Rodrigo Ramele, Ana Julia Villar and Juan Miguel Santos}

@@

% Contact information of the corresponding author
\corres{Correspondence: rramele@itba.edu.ar; Tel.: +54-11-2150-4800(5834)}

@@

% Abstract (Do not use inserted blank lines, i.e. \\) 
\abstract{The analysis of Electroencephalographic (EEG) signals is of ulterior importance to elucidate patterns that could improve the implementation of Brain Computer Interfaces (BCI). These systems are meant to provide alternative pathways to transmit volitional information which could potentially enhance the quality of life of patients affected by neurodegenerative disorders or improve Human Computer Interaction systems.  Of particular interests are those which are based on the recognition of Event-Related Potentials (ERP) because they can be elicited by external stimuli and used to implement spellers, to control external devices or even avatars in virtual reality environments.  This work mimics what electroencephalographers have been doing clinically, visually inspecting and categorizing phenomena within the EEG by the extraction of features from the images of the plots of the signals.  It also aims to provide a framework to analyze, characterize and classify EEG signals, with a focus on the P300, an ERP elicited by the oddball paradigm of rare events.  The validity of the method is shown by offline processing a public dataset of Amyotrophic Lateral Sclerosis (ALS) patients.}

@@

Although recent advances in neuroimagining techniques (particularly radio-nuclear and radiological scanning methods) \citep{Schomer2010} have diminished the prospects of the traditional Electroencephalography (EEG), the advent and development of digitalized devices has pressed for a revamping of this hundred years old technology.  Their versatility, ease of use, temporal resolution, ease of development and fabrication, and its proliferation as consumer devices, are pushing EEG to become the de-facto non invasive portable or ambulatory method to access and harness brain information \cite{DeVos2014}

A key contribution to this expansion has been the field of Brain Computer Interfaces (BCI) \citep{WolpawJonathanR2012} which is the pursuit of the development of a new channel of communication particularly aimed to persons affected by neurodegenerative diseases.

One noteworthy aspect of this novel communication channel is the ability to volitionally transmit information from the Central Nervous System (CNS) to a computer device and from there use that information to control a wheelchair \citep{Carlson2013}, as input to a speller application \citep{Guger2009a}, in a Virtual Reality environment \citep{Lotte2013} or as aiding tool in a rehabilitation procedure \citep{Jure2016}.  The holly grail of BCI is to implement a new complete and alternative pathway to restore lost locomotion \citep{WolpawJonathanR2012}.

EEG signals are remarkably complex and have been characterized as a multichannel non-stationary stochastic process.  Additionally, they have high variability between different subjects and even between different moments for the same subject, requiring adaptive and co-adaptive calibration and learning procedures \citep{Clerc}.  Hence, this imposes an outstanding challenge that is necessary to overcome in order to extract information from raw EEG signals.

Moreover, EEG markers \citep{Clerc} that can be used to volitionally transmit information are limited, and each one of them has a particular combination of appropriate methods to decode them. Inevitably, it is necessary to implement many distinct and specialized algorithmic methods, to filter the signal, enhance its SNR, and try to determine some meaning out of it.

%Patients suffering from ALS their brain signals do suffer slow and deteriorating changes that may lead to different EEG patterns alterations \citep{Nijboer2009,Riener2014}.

%Besides novel applications of BCI, the goal of the entire discipline is to provide communication assistance to people affected by neuro-degenerative diseases.  It is understood that for people suffering from ALS, even though this pathology do not harm the CNS, their brain signals do suffer slow and deteriorating changes that may lead to different EEG patterns alterations \citep{Nijboer2009,Riener2014}.

BCI has gained mainstream public awareness with worldwide challenge competitions like Cybathlon  \citep{Riener2014} and even been broadcasted during the inauguration ceremony of the 2014 Soccer World Cup.  New developments have overcome the out-of-the-lab high-bar and they are starting to be used in real world environments \citep{Huggins2016}.  However, they still lack the necessary robustness, and its performance is well behind any other method of human computer interaction, including any kind of detection of residual muscular movement \citep{Clerc}.

In this work, a new method to characterize and classify EEG signals is presented, expanded and detailed.  Its validity is verified by processing offline data for ALS patients.  This is the continuation of the work previously presented in \citep{Ramele2016}, where it was applied to rhythmic patterns, and that it can be extended to describe transient events like those produced by P300 Event Related Potential \citep{Knuth2006}.

The method is based on the morphological analysis of the shape of the EEG signal \citep{Alvarado-Gonzalez2016,Yamaguchi2009} and was inspired by mimicking what traditionally electroencephalographers have been performing for almost a century: visually inspecting  raw signal plots \citep{Hartman2005}.

This paper reports a method to, (1) characterize EEG signals based on the identification of their structure in the shape domain using histograms of oriented gradients extracted from signal plots, and (2) how this characterization can be used to implement a BCI classification scheme to identify Event Related Potentials, particularly the well-known P300, on an offline and public dataset.

This article unfolds as follows: in Section \ref{Feature} the Feature Extraction based on Histogram of Gradients of the Signal Plot method is explained: section \ref{Preprocessing} and \ref{Pipeline} describe the processing pipeline.  Section \ref{Plot}  clarifies how the image of the signal plot is constructed whereas Section \ref{SIFT}  describes in detail the feature extraction procedure.  Section \ref{Classification}  presents the classification algorithm based on Naive Bayes Near Neighbor  and the final Section shows results and discussion where we expose our remarks, conclusions and future work.

@@

\subsection{Feature Extraction based on Histogram of Gradients of the Signal Plot} \label{Feature}

The P300 \citep{Farwell1988,Knuth2006} is a positive deflection of the EEG signal which occurs 300 ms after the onset of a rare and deviant stimulus that the subject is expected to attend.  It is produced under the oddball paradigm \cite{WolpawJonathanR2012} and though it is quite consistent across different subjects, it has a lower amplitude ( $ \pm 5 \mu V $ ) compared to basal EEG activity, reaching a SNR of around -15 db estimated based on the amplitude of the P300 response signal divided by the standard deviation of the background EEG activity \citep{Hu2010}.  This signal can be used to implement a speller application by means of a Speller Matrix \citep{Farwell1988} (Fig. \ref{fig:p300matrix}) where the flashings of rows and columns provide the deviant stimulus required to elicit this physiological response.   Each time a row or column that contains the desired letter flashes, the corresponding synchronized EEG signal should also contain the P300 signature and by detecting it, the selected letter can be identified.

\begin{figure}[H]
\centering
\includegraphics[width=10cm]{openvibep300matrix.png}
\caption{The P300 Matrix\citep{Farwell1988} used in the OpenVibe Open Source software \citep{Renard2010}. Rows are labeled from 1 to 6 and columns from 7 to 12.  Rows/columns flash intermittently in random permutations.}
\label{fig:p300matrix}
\end{figure}

%The proposed algorithm was designed to perform specifically this decoding.

% of the selected letter based on the detection of the P300 Event Related Potential response on the corresponding EEG signal.

\subsubsection{Preprocessing} \label{Preprocessing}

The first step consists of the enhancement of the SNR of the P300 pattern above the level of basal EEG. 
The processing pipeline starts by applying a notch filter to the raw signal, a 
4th degree 10 Hz lowpass Butterworth filter and finally a decimation with an FIR filter of order 30 from the original 256 Hz down to 16 Hz \citep{Krusienski2006}.

\subsubsection{Processing Pipeline} \label{Pipeline}

\begin{itemize}
\item \textbf{Artefact Removal}: The EEG signal matrix is processed on a channel by channel basis.   For every 12 flashing stimuli, i.e. one complete sequence of intensification of each of the 6 rows plus the 6 columns, a basic artefact elimination procedure is implemented by removing the entire segment when any signal deviates above/bellow $ \pm 70 \mu V $.
\item \textbf{Segmentation}: For each of the 12 stimuli,  a window of 1 second of the multichannel signal is extracted, starting from the stimulus onset, corresponding to each row/column intensification.  Two of these segments are labeled as \textit{hit}, whereas the remaining 10 are labeled as \textit{no hit}.  A hit represents that the EEG segment should contain the ERP signature time-locked to the flashing stimulus. 
\item \textbf{Signal Averaging}: The P300 ERP is deeply buried under background EEG so the traditional approach to identify it is by point-to-point averaging the time-locked stacked signal segments.  Hence the values which are not related to, and not time-locked to the onset of the stimulus are canceled out \citep{Liang2008}. 
\end{itemize}

This last step determines the operation of any P300 Speller.  In order to obtain an improved signal in terms of its SNR, repetitions of the sequence of row/column intensification are necessary.  And, at the same time, as long as more repetitions are needed, the ability to transfer information faster is diminished, so there is a trade-off that must be acutely determined.

\subsubsection{Signal Plotting} \label{Plot}

The underlying idea of this method is to generate a template for the signal,  based on the image plot.  Hence, the first step is its transformation into a temporary binary image.

The signal is first scaled and standardized (i.e. z-score) by

\begin{equation}
\tilde{x}(t,c) = \left \lfloor{ \gamma \cdot \frac{( x(t,c) - \bar{x}(c)  )}{ \sigma_x(c) } }\right \rfloor
\end{equation}

\noindent where $\gamma$ is the image scale, $t$ is the time and $ x(t,c) $ is the point-to-point averaged EEG matrix defined for each $t$ and for a channel $c$. Lastly, $\bar{x}(c)$ and $ \sigma_x $ are the mean and standard deviation of $x$.

Consequently, the image is constructed by placing the sample points according to

\begin{equation}
I(z_1,z_2) = \left\{ \begin{array}{rl}
255 & \text{if} \,  z_1 = \gamma \cdot t; z_2 = \tilde{x}(t,c) + z(c) \\
0   & \mbox{otherwise}
\end{array}\right.
\label{eq:images}
\end{equation}

\noindent where $ z_1$ and $z_2$ iterate over the width (based on the length of the signal segment) and height (based on the peak-to-peak amplitude) of the newly created image.  The function $z(c)$ is the \textit{zerolevel} which is the location on the image where the signal's zero value should be located in order to fit the entire signal within the image:

\begin{equation}
z(c) = \left \lfloor{ \frac{\max \tilde{x}(t,c)  - \min \tilde{x}(t,c) }{2} }\right \rfloor -   \left \lfloor{ \frac{\max \tilde{x}(t,c)  + \min \tilde{x}(t,c)}{ 2} }\right \rfloor
\label{eq:zerolevel}
\end{equation}

In order to complete the plot from the pixels, the Bresenham \citep{Ramele2016} algorithm is used to interpolate straight lines between each pair of these consecutive pixels.

%\noindent and is used to derive local representations that captures the visual shape of the signal.

\subsubsection{Feature Extraction: Histogram of Oriented Gradients} \label{SIFT}

%is calculated on local image patches centered on sample points \citep{Vedaldi2010}.

On the generated image, a keypoint is placed on a precise location along the signal trace over the image plot (Fig. \ref{fig:sampledescriptor}).  Around that keypoint, a local image patch is constructed as follows:  it is first divided in 16 blocks of size $ 3 \cdot s $ each, arranged in a 4x4 grid and centered on $T = (x,y)$.

A local representation of the shape of the signal within the patch can be described by obtaining the gradient orientations on each of the 16 blocks and creating a histogram of the oriented gradients.  This technique is based on Lowe's SIFT \citep{Lowe2004} descriptor, and it is biomimetically inspired in how the visual cortex detects shapes by analyzing orientations.   In order to calculate this histogram, the interval $ [0-360]$ of possible angles is divided in 8 bins, each one at 45 degrees.

The values of the image gradients for the 8 angle bins $ p = {0, 45, 90, 135, 180, 225, 270, 315} $ and for each grid block bin $ i,j = {1,2,3,4} $  are accumulated in the 3-dimensional histogram $ h $ through

\begin{equation}
 h(p,i,j) = 3 \cdot s \int w_\mathrm{ang}(\angle J(\mathbf{x}) - \theta_p)\, w_{ij}\left(\frac{\mathbf{x} - T}{3 s}\right)\, |J(\mathbf{x})|\, d\mathbf{x}\, d\theta_p
\label{eq:histogram}
\end{equation}

\noindent where $s$ is the size of the local patch, which can be converted into pixels by doing $ Px = 4 \cdot 3 \cdot s $ ,  $ |J(\mathbf{x})| $ is the norm of the gradient vector found at each one of the 16 blocks of the patch, whereas $\angle J(\mathbf{x}) $ is the angle of the gradient vector and  $\theta_p$ is the angle bin.  On the other hand $ \textbf{x} = (x_i, x_j) $ corresponds to each one of the pixels of the local patch, and $ w_\mathrm{ang}  $  and $ w_{ij} $ are linear interpolation functions \citep{Lowe2004,Vedaldi2010}.  Lastly, the fixed value of $ 3 $ is a magnification factor which corresponds to the number of pixels per each grid block when the size of the patch is the unity.  As the patch is 4x4, it gives a descriptor of 128 dimension as shown on Figure \ref{fig:sampledescriptor}.

\begin{figure}[H]
\centering
\includegraphics[width=6cm]{sampledescriptor.png}
\includegraphics[width=9cm]{histogramsbines.png}
\caption{(Left) The local patch is located on a precise location within the generated plot. The 4x4 block grid can be seen on the patch. (Right) On each grid block, eight bins are used to describe the eight different orientations, and they are later arranged in a 128-dimension vector (4x4x8).}
\label{fig:sampledescriptor}
\end{figure}

\subsubsection{Classification} \label{Classification}

The classification is carried out by using a discriminative semi-supervised classification method, based on the Naive Bayes Nearest Neighbor (NBNN) \citep{Boiman2008}.

%The first step consists in extracting labeled descriptors from the training set for the  which are thus grouped together in two KD-tree \citep{Vedaldi2010} database structures. and group those marked as \textit{hit} in a template set $ T $

@@

While doing classification based on local features, which encode partial information, one problem that frequently arises is how to ensemble the individual classification of each local characteristic to the image that were used to derive those descriptors.

The NBNN algorithm tackles this problem by comparing each image against a whole classification class which is characterized by the set of descriptors that are closest to each one of the descriptors obtained for the query image.  The algorithm is based on

\begin{equation}
\hat{C} = \arg \min_C \sum_{}^{} \left\lVert d_i - NN_C(d_i) \right\rVert ^2
\label{eq:classification}
\end{equation}

\noindent where the predicted class $\hat{C}$ of a query image is calculated as the class $ C $ that minimize the summation of the $ L^2 $ distance between each descriptor $ d_i $ that belongs to the query image and its corresponding near neighbour $ NN_C(d_i) $ descriptor for class C.

This basic method was adapted to the P300 Speller and the classifier works according to the following modified equation:

\begin{equation}
\hat{Q}_{r\,c} = \arg \min_q \sum_{x_i \epsilon NN_T^k(q)}^{} \left\lVert x_i - q \right\rVert ^2
\label{eq:multiclassification}
\end{equation}

The first step consists in extracting labeled \textit{hit} descriptors from the training set which are thus grouped together in a KD-tree \citep{Vedaldi2010} template set $ T $.  On decoding stage, 12 new images from the signal segments are generated, and their descriptors extracted.  The 12 descriptors are divided in rows and columns.  For each one of the 6 query descriptors $ q $ for row/column, its $ k $ nearest neighbours in the template set $T$, $ NN_T^k(q) $ (acquired during the training phase)  are obtained and the distance between each one of them, $ x_i $, and the query descriptor $ q $ is summarized.  The query descriptor $ q $ that minimizes this summation is the one that is chosen, first between the 6 rows, and then between the remaining 6 columns.  By performing this procedure, the correct letter can be identified by matching the row $ r $ and column $ c $ from the P300 speller matrix.

Especially in the case of the P300 response, the oddball paradigm requires that one of the stimuli need to be infrequent so that will unavoidably force the data to be unbalanced \citep{Tibon2015}.  The NBNN method suffers from biased classification on unbalanced classes \citep{Fornoni2014}.  By reversing the roles of the query and the class in Eq. \ref{eq:multiclassification}, it is only necessary to obtain the template set $ T $ with the learned descriptors representative of the P300 ERP, hence avoiding the problem of unbalanced classes.

%Additionally,

%El tema del balanceo lo mencionas y marcas lo cual está muy bien, pero no decí cómo lo solucionaste. Si llegaras a tener un poco de tiempo podrías mencionarlo.

\subsection{Experimental Protocol} \label{Protocol}

To verify the validity of the proposed framework and method, the public dataset 008-2014  \citep{Riccio2013} published on the BNCI-Horizon website \citep{Brunner2014} by  IRCCS Fondazione Santa Lucia, was used to perform an offline BCI Simulation to decode the spelled words from the provided signals.  The algorithm was implemented using  VLFeat  \citep{Vedaldi2010} Computer Vision libraries on MATLAB 2014a (Mathworks Inc., Natick, MA, USA).

\subsubsection{P300 ALS Public Dataset}

The experimental protocol used to generate this dataset is explained in \citep{Riccio2013} but can be summarized as follows:  8 subjects with confirmed diagnoses but on different stages of ALS disease, were recruited and accepted to perform the experiments. The P300 detection task designed for this experiment consisted of spelling 7 words (runs) of 5 letter each, using the traditional P300 Speller Matrix \citep{Farwell1988} where the flashings of rows and columns provide the deviant stimulus required to elicit this physiological response.  The first 3 runs were used for training and the remaining 4 for testing with visual feedback.  A trial, as defined by the BCI2000 platform \citep{Schalk2004}, was every attempt to select a letter from the speller, and it was composed of signal segments corresponding to 10 repetitions of flashes of 6 rows and 6 columns of the traditional 6x6 P300 matrix , yielding 120 repetitions.  Flashing of row/columns was performed for 0.125 s, following by a resting period (i.e. Inter-stimulus interval) of the same length.  After 120 repetitions an inter-trial pause was included before resuming with the following letter.

The recorded dataset was sampled at 256 Hz and consisted of an EEG matrix for electrode channels Fz,Cz,Pz,Oz,P3,P4,PO7 and PO8, identified according to the 10-20 International System,  for each one of the 8 subjects.

In order to asses and verify the identification of the P300 response, subjects are instructed to perform a copy-spelling task where they have to fix their attention to successive letters in order to copy a previously determined set of words (in contrast to an online free-running operation of the speller where each users decides on its own what letter to choose).

\subsubsection{Parameters}

Parameters were selected according to the experimental protocol. As the P300 event latency and amplitude vary greatly between subjects, it is necessary to provide a patch that will be able to capture an entire transient event.  Equations \ref{eq:mapping1} and \ref{eq:mapping2} can be used to map the original signal parameters to local image patch structure.

\begin{equation}
s = \frac{\Delta \mu V}{4 \cdot 3} \cdot \gamma 
\label{eq:mapping1}
\end{equation}

\begin{equation}
s = \frac{\lambda \cdot Fs}{4 \cdot 3} \cdot \gamma
\label{eq:mapping2}
\end{equation}

\noindent where $ Fs $ is the sample frequency of the EEG signal (downsampled to 16 Hz), $ \lambda $ is the length in seconds covered by the patch, and $\Delta  \mu V $ corresponds to the amplitude in microvolts that can be covered by the height of the patch. By using $ s = 3 $ and a quadruple scale of the image $ \gamma = 4 $ this gives the local patch, and the descriptor, the ability to identify events of 9 $ \mu V $ of amplitude, a resolution of $ 1 $ Pixel $ = \frac{1}{4} \mu V $ and span   of $ \lambda = 0.56 s $ (because the patch is a geometric square, $ s $ must be the same to map both the height and the required span of the signal).  Finally, descriptor locations $ T $ were selected at $ x =  0.55 s \cdot Fs \cdot \gamma = 35 $ and $ y = z(c) $ (Eq. \ref{eq:zerolevel}).

%\begin{figure}[H]
%\centering
%\includegraphics[width=16cm]{localpatch.png}
%\caption{The local patch is mapped to the original image by their correspondence .}
%\label{fig:sampledescriptor}
%\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results and Discussion} \label{Results}
\label{section:results}
In Figure \ref{fig:subjectaveraged} the grand average (point-to-point) for all the subjects using the information from all the segments can be shown.  The P300 characteristic curve can be seen particularly in subjects 2, and 6 and in a lesser extend in the remaining subjects. In order to correctly decode the selected letter from each trial, particular care was observed to avoid unbalanced number of epochs  (i.e. an unequal number of epochs on each condition), because that may introduce a bias in the classification procedure (the variance of averaged signals is inversely proportional to the number of samples and the procedure would be discriminating signals with different variances).

>>>>

%=================================================================
% Full title of the paper (Capitalized)
\Title{EEG characterization and classification based on histograms of visual gradients}

% Authors, for the paper (add full first names)
\Author{Rodrigo Ramele $^{1,\dagger}$, Ana Julia Villar $^{1}$ and Juan Miguel Santos  $^{1}$*}
% Authors, for metadata in PDF
\AuthorNames{Rodrigo Ramele, Ana Julia Villar and Juan Miguel Santos}

@@

% Contact information of the corresponding author
\corres{Correspondence: rramele@itba.edu.ar; Tel.: +54-9-11-4193-9382}

@@

% Abstract (Do not use inserted blank lines, i.e. \\) 
\abstract{The analysis of Electroencephalograpy (EEG) signals for Amyotrophic Lateral Sclerosis (ALS) patients is of ulterior importance to elucidate different patterns or markers that could be potentially used to aid in diagnosis or treatment and to implement Brain Computer Interface (BCI) systems which may provide alternative pathways to transmit volitional information to enhance patient's quality of life.  Of particular interests are those which are based on the recognition of Event-Related Potentials (ERP) which are shown to be more robust to the unavoidable physical impairment which comes with this motor neuron disease.  This work mimics what electroencephalographers has been doing clinically, visually inspecting and categorizing phenomena within the EEG, and provides a framework to analyse and characterize EEG signals, with a focus on the P300, an ERP elicited by the oddball paradigm of rare events.  The validity of the method is shown by off-line processing a public dataset of ALS patients.}

@@

Although impressive advances in neuroimagining techniques (particularly radio-nuclear and radiological scanning methods) \citep{Schomer2010} has diminished the prospects of the traditional Electroencephalography (EEG), the advent and development of digitalized devices has pressed for a revamping of this hundredth years old technology.  Their versatility, ease of use, temporal resolution, ease of development and fabrication, and its proliferation as consumer devices, is pushing EEG to become the de-facto non invasive method to access and harness brain information.

Key contribution to this development has been the field of Brain Computer Interfaces (BCI) \citep{WolpawJonathanR2012} which is the pursuit of the development of a new channel of communication particularly aimed to persons affected by neuro-degenerative diseases.

One very noteworthy aspect of this communication channel is the ability to volitionally transmit information from the Central Nervous System (CNS), to a computer device and from there use that information to control a wheelchair \citep{Carlson2013}, as input to a speller application \citep{Guger2009a}, or as aiding tool in a rehabilitation procedure \citep{Jure2016}.  The holly grail of BCI is to implement a new complete and alternative pathway to restore lost locomotion \citep{WolpawJonathanR2012}.

EEG signals are remarkably complex and have been characterized as a multichannel non-stationary stochastic process and even considered random enough as to be used as a source of pseudo random number generator \citep{Chen2014}. Additionally, they have a high variability between different subjects and even between different moments for the same subject, requiring an adaptive and co-adaptive calibration and learning procedures \citep{Clerc}.  Hence, this imposes an outstanding challenge that is necessary to overcome in order to extract information from raw EEG signals.

EEG markers \citep{Clerc} that can be used to transmit volitionally information are limited, and each one of them has a particular set or combination of appropriate methods to decode them. Inevitably, it is often necessary to implement many distinct and specialized algorithmic methods, to filter the signal, enhance its SNR, and try to determine some meaning out of it.

In this work a new method to characterize EEG signals is presented, expanded and detailed.  Its validity is verified by processing off-line data for ALS patients.  This is the continuation of the work previously presented in \citep{Ramele2016}, where it was applied to rhythmic patterns, and that it can be extended to describe transient events like those produced by the P300\citep{Knuth2006}.

The method is based on the morphological analysis of the shape of the EEG signal\citep{Alvarado-Gonzalez2016,Yamaguchi2009} and was inspired by mimicking what traditionally electroencephalographers have been performing for almost a century: visually inspecting  raw signals.

%This paper reports on a study done to address
%the hypothesis that communication task complexity
%(copy-spelling vs. novel text generation of a picture
%description) is correlated with BCI performance accuracy
%and P300 characteristics.

In brief, this paper reports a straightforward method to, first characterize EEG signals based on the identification of their morphological shapes, and how this characterization can be used to implement a BCI scheme to identify Event Related Potentials, particularly the well-known P300, on an off-line and public dataset.

Besides new novel applications of BCI, the goal of the entire discipline is to provide communication assistance to people affected by neuro-degenerative diseases.  It is understood that for people suffering from ALS, even though this pathology do not harm the CNS, their brain signals do suffer slow and deteriorating changes that may lead to different EEG patterns \citep{Nijboer2009,Riener2014}.

BCI has gained mainstream public awareness and recently even held a peculiar Olympics \citep{Riener2014} and been broadcasted during the inauguration of the last World Cup.  New developments have overcome the out-of-the-lab high-bar and they are starting to be used in real world environments\citep{Huggins2016}.  However, they still lack the necessary robustness, and its performance is well behind any other method of human computer interaction, including any kind of detection of remaining muscular movement \citep{Clerc}.

To verify the validity of the proposed framework and method, the public dataset 008-2014 published on the BNCI-Horizon website \citep{Riccio2013} was used to perform a binary classification task on the provided signals.  The algorithm was implemented using the VLFeat \citep{Vedaldi2010} Computer Vision libraries on MATLAB 2014a (Mathworks Inc., Natick, MA, USA).

\subsection{Experimental protocol}

The protocol is explained in \citep{Riccio2013} but can be summarized as follows:  8 subjects with confirmed diagnoses but on different stages of ALS disease, were recruited and accepted to perform the experiments. The P300 task objective was meant to spell 7 words of 5 letter each.  The first 3 runs were used for training and the remaining 4 for testing with visual feedback.  A trial, for this experiment, was defined as every attempt to select a letter from the speller, and it was composed of 10 repetitions of flashes of 6 rows and 6 columns of the traditional 6x6 P300 matrix , yielding 120 repetitions.  Flashing of row/columns was performed for 0.125 s, following by a resting period of the same size.  After 120 repetions and inter-trial pause was included before resuming with the following letter.

The gathered dataset was sampled at 256 Hz and consisted of EEG matrix for electrode channels Fz,Cz,Pz,Oz,P3,P4,PO7 and PO8, identified according to the 10-20 International System,  for each one of the 8 subjects.  This dataset was published in the popular BNCI-Horizon 2020 website \citep{Brunner2014}.

@@

\subsection{Algorithm}

\subsubsection{Preprocessing}

The mandatory first step is to help to enhance the SNR of the P300 pattern above the level of basal EEG. The processing pipeline starts with applying a notch filter to the raw signal, a decimation with a FIR filter of order 30 from the original 256 Hz to 32 Hz and finally applying a 4th degree 10 Hz lowpass Butterworth filter.

\subsubsection{Segmentation, Artefact Removal and Signal Averaging}

The EEG matrix is processed on a channel by channel basis.  For each word, each letter, a segment of EEG signal is extracted that encompass 10 repetitions of 12 flashings, 6 rows and 6 columns.  For every 12 stimuli flashing, i.e. one complete sequence of repetitions of each row/column, a basic artefact removal procedure of removing the entire segment if any signal goes above 70 $\mu$V is applied.

From every segment of 12 flashings, two epoch corresponding to a hit and two matching a no-hit are constructed as segments of 1s length (256 sample points).

These segments, 20 for each class, are later point-to-point averaged for the whole 120 repetitions group.  This finally lead to 2 averaged signals for each of the 5 letters, for the 7 words, comprising a set of 70 averaged signals\citep{Liang2008}.

\subsubsection{Features}

The central part of this method is the feature generation: the histogram of visual gradients.  To do so, the first step is the transformation of the signal into a temporary binary image.

\begin{equation}
\tilde{x}(t,c) = \left \lfloor{ \gamma \cdot ( x(t,c) - \bar{x}(t,c)  )}\right \rfloor
\end{equation}

where $\gamma$ is the image scale, $t$ is time and $ x(t,c) $ is the EEG matrix defined for each $t$ and for a particular channel $c$.

\begin{equation}
I(z_1,z_2) = \left\{ \begin{array}{rl}
255 & z_1 = \gamma \cdot t; z_2 = \tilde{x}(t,c) + z(c) \\
0   & \mbox{otherwise}
\end{array}\right.
\label{eq:images}
\end{equation}

From the Equation \ref{eq:images} the image $ I(z_1,z_2) $ is constructed an used to derive local representations that will capture the visual shape of the signal.

Finally, the histogram of gradients, which is based on a very successful technique from Computer Vision, the SIFT \citep{Lowe2004} descriptor, is calculated on local image patches centered on sample points \citep{Vedaldi2010}.

\begin{equation}
 h(p,i,j) = 3 \cdot s \int w_\mathrm{ang}(\angle J(\mathbf{x}) - \theta_p)\, w_{ij}\left(\frac{\mathbf{x} - T}{3 s}\right)\, |J(\mathbf{x})|\, d\mathbf{x}
\label{eq:histogram}
\end{equation}

where $s$ is the size of the local patch, $\angle J(\mathbf{x})$ is the gradient found at each of the 16 blocks of the patch aligned with the angle $\theta_p$.  The patch is divided in 16 blocks, arranged in a 4x4 grid and centered on $T = (x,y)$, a location within the image.  On each block, the bin consists of 45 degrees obtaining 8 aligning angles. The Equation \ref{eq:histogram} accumulates the values of the image gradients aligned with each angle bin.  This gives a descriptor of 128 dimension as shown on Figure \ref{fig:sampledescriptor}.

\begin{figure}[H]
\centering
\includegraphics[width=16cm]{sampledescriptor.eps}
\caption{The local patch is located around a sample point plotted on the temporary signal. All the sample points are interpolated using the Bresenham algorithm. The 4x4 grid can be seen on the descriptor and also arrows showing the dominant gradient on each block.}
\label{fig:sampledescriptor}
\end{figure}

\subsubsection{Classification}

The classification is carried out by using the discriminative semi-supervised classification method, the Naive Bayes Nearest Neighbout (NBNN) \citep{Boiman2008}. % algorithm which can be described as follows.

\begin{equation}
\hat{C} = argmin_C \sum_{}^{} \left\lVert d_i - NN_C(d_i) \right\rVert ^2
\end{equation}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}

In Figure \ref{fig:subjectaveraged} the grand average (point-to-point) for all the subjects is shown.  The P300 characteristic curve can be seen particularly in subjects 2, and 6 and in a lesser extend in the remaining subjects. In order to obtain a valid binary classification on averaged signals, particular care was observed to avoid unbalanced samples, because that may introduce a bias in the classification procedure (signal variance is proportional to the square root of the number of samples and the procedure would be discriminating signals with different variances).Particularly in the case of the P300 response, the oddball paradigm requires that one of the stimulus need to be infrequent so that will unavoidably force that the data will be unbalanced (that is, an unequal number of trials in each condition).

\begin{figure}[H]
\centering
\includegraphics[width=16cm]{subjectaveraged.eps}
\caption{Point-To-Point grand averages of epochs obtained for hits (solid line) and no hits (dashed line) for each one of the 8 subjects for channel Cz. The P300 characteristic curve can be well identified on subjects 2 and 6.}
\label{fig:subjectaveraged}
\end{figure}

Results are shown in Table \ref{tab:results} where BCI accuracy
for the Cz channel is calculated and compared with the values obtained by the published of the dataset.  Additionally the best performing channel is informed and its accuracy value.  It is of particular interest that using this method, the best performing channel was not always Cz, and instead occipital channels PO8 and PO7 showed very good performance indeed \citep{Huggins2016,Jure2016}.

% (CITE TAYLOR AND  FRANCOS PAPER).

\begin{table}[H]
\caption{Results obtained for Accuracy levels by 3-fold cross validation. The values reported by the dataset publishers for Cz are reproduced here for comparison. Additionally, channel Cz BCI accuracies can be seen as well as the performance levels obtained for the BPC, the best performing channel for each subject.}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccccc}
\toprule
\textbf{Participants}	& \textbf{Original Cz}	& \textbf{ACC at Cz}	& \textbf{BPC}	& \textbf{Performance}\\
\midrule
1     &   0.84 &   0.79 & Cz  &   0.79 $\pm$ 0.01 \\
2     &   0.86 &   0.81 & PO7 &   0.93 $\pm$ 0.01 \\
3     &   0.87 &   0.78 & Cz  &   0.78 $\pm$ 0.03 \\
4     &   0.86 &   0.68 & PO8 &   0.90 $\pm$ 0.01 \\
5     &   0.86 &   0.80 & PO7 &   0.93 $\pm$ 0.01 \\
6     &   0.89 &   0.96 & Cz  &   0.96 $\pm$ 0.01 \\
7     &   0.89 &   0.78 & PO7 &   0.93 $\pm$ 0.01 \\
8     &   0.92 &   0.91 & PO7 &   1.00 $\pm$ 0.00 \\
\bottomrule
\end{tabular}
\label{tab:results}
\end{table}

The ITR, or BTR, in the case of reactive BCIs \citep{WolpawJonathanR2012} strongly depends on the amount of signal averaging required to transmit a valid and robust selection.  There is a trade-off that needs to be balanced between the required number of repetitions for each trial to guarantee robust transmission and the the achieved speed of transmission affected by the repetions.  As detailed in the table \ref{tab:singletrialreduction} this method is only reduced in a 20 percent in accuracy levels even by using only one sequence of repetitions.

\begin{table}[H]
\caption{Results obtained for Accuracy levels by performing a BCI simulation for the testing dataset. The BPC, best performing channel is shown and also performance levels for the 120, 60 and 12 repetitions. Finally the PR, performance reduction is described.  Almost all the subjects performed around 70\% even with by averaging 2 signals for each class.}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{cccccc}
\toprule
\textbf{Participants}	& \textbf{BPC}	& \textbf{ACC@120}	& \textbf{ACC@60}	&  \textbf{ACC@12} & \textbf{PR}\\
\midrule
1 & Cz &   0.80 &   0.82 &   0.72 & 9\% \\
2 & PO7 &   0.93 &   0.78 &   0.75 & 18\%\\
3 & Cz &   0.80 &   0.82 &   0.72 & 9\%\\
4 & PO8 &   0.93 &   0.78 &   0.65 & 29\%\\
5 & PO7 &   0.93 &   0.78 &   0.75 & 18\%\\
6 & Cz &   0.80 &   0.82 &   0.72 & 9\%\\
7 & PO7 &   0.93 &   0.78 &   0.75 & 18\%\\
8 & PO7 &   0.93 &   0.78 &   0.75 & 18\%\\
\bottomrule
\end{tabular}
\label{tab:singletrialreduction}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Discussion}
%
%This method, different from other methods which is based on the nonlinearity of the gradient of histograms which can be used to detect 
%
%is also based on how the image look like.
%
%SNR of p300 and how to detect it
%
%Check if you can use this to detect any kind of transient signal.
%
%Compare if it is possible with the descriptors from one subject, discriminate the others.
%
%Channal identification based on the metric distance between the bags
